---
title: "smt3 KD analysis genes"
# knit: (function(input_file, encoding) {
#  out_dir <- 'docs';
#  rmarkdown::render(input_file,
# encoding=encoding,
# output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
# author: "Maria Ninova"
# date: "April 20, 2020"
# output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sleuth)
library(dplyr)
library(stringr)
library(ggplot2)
library(viridis)
library(DT)
units<-"scaled_reads_per_base"
#units<-"est_counts" # if doing transcript-centered.
```


This is gene level analysis, using gene_mode=T. ?sleuth_prep
gene_mode: Set this to TRUE to get the old counts-aggregation method for doing gene-level analysis. This requires aggregation_column to be set. If TRUE, this will override the p-value aggregation mode, but will allow for gene-centric modeling, plotting, and results.

## Data pre-processing

The original FASTQ files (50 bp, single-end, reverse stranded) were processed with kallisto using a custom transcriptome, which is a combination of the REFSEQ mRNA sequences AND RepBase D. melanogaster TE seqeunces (refGene_plus_repbase)

> ./kallisto quant --single -t 4 -l 200 -s 50 -b 30 --rf-stranded -i transcriptome_plus_repbase/refGene_plus_repbase -o YL25 YL25_shW_RNAseq_1st.fq.gz

> ./kallisto quant --single -t 4 -l 200 -s 50 -b 30 --rf-stranded -i transcriptome_plus_repbase/refGene_plus_repbase -o YL26 YL26_shSmt3_RNAseq_1st.fq.gz

> ./kallisto quant --single -t 4 -l 200 -s 50 -b 30 --rf-stranded -i transcriptome_plus_repbase/refGene_plus_repbase -o YL27 YL27_shW_RNAseq_2nd.fq.gz

> ./kallisto quant --single -t 4 -l 200 -s 50 -b 30 --rf-stranded -i transcriptome_plus_repbase/refGene_plus_repbase -o YL28 YL28_shSmt3_RNAseq_2nd.fq.gz

Next, we use these files for sleuth analysis, aggregating the data per gene. In order to aggregate, we need a dictionary of transcript to gene.
It is possible to do transcript-centric analysis as an alternative, but it doesn't change the results that R1 and R2 stand out as most strongly affected. 

```{r create_so, warning=FALSE}

s2c<-data.frame(sample=c("YL25","YL27","YL26","YL28"),condition=c("ctrl", "ctrl","KD","KD"), 
                path=c("../kallisto/YL25/","../kallisto/YL27/", "../kallisto/YL26/","../kallisto/YL28/"))
s2c$path<-as.character(s2c$path)
print(s2c)

#this is to create a transcript to gene name dictionary - I just transform each fasta header.
t<-read.table("../kallisto/YL26/abundance.tsv", header=T, as.is=T) 
t2g<-data.frame("target_id"=t$target_id, 
                     gene=sub(":CDS.*","",sub(".*gene=","", t$target_id)))
t2g$coding<-grepl("CDS", t2g$target_id)
t2g$TE<-grepl("\\|", t2g$target_id)

# number of elements of different type: with CDS annotation, RepBase entries (TE), and none (ncRNAs or genes with no CDS)
t2g %>% select(gene,coding, TE) %>% distinct() %>% group_by(coding, TE) %>% tally()

# make "so"
so <- sleuth_prep(s2c, ~condition, target_mapping = t2g, extra_bootstrap_summary=T, read_bootstrap_tpm = TRUE, aggregation_column = 'gene', gene_mode=TRUE)

```

***
***

#### Input values

Here, we can inspect the input data table.

```{r, display_matrix, message=FALSE}
units

table1<-sleuth_to_matrix(so, which_df = "obs_norm", which_units = units) %>% as.data.frame() %>% 
  rename(shW.1=YL25, shW.2=YL27, shSmt3.1=YL26, shSmt3.2=YL28) 

table1 %>% round(digits=2) %>% datatable(options=list(pageLength=5))

```


***
***

## Analysеs

Running sleuth Likelihood Ratio test (LRT), and displaying a table of significantly changed genes, qval<0.05. This table DOES NOT have information about up- or down- regulation.


```{r sleuth_lrt,  message=FALSE}

so <- sleuth_fit(so, ~condition,'full')
so <- sleuth_fit(so, ~1, 'reduced')
so <- sleuth_lrt(so, 'reduced', 'full')

sleuth_table <- sleuth_results(so, 'reduced:full', test_type = 'lrt', show_all = TRUE)

dplyr::filter(sleuth_table, qval <= 0.05) %>% datatable(options = list(pageLength=5))

```

***
***

Running wald test & plotting MA plot from these analysis (note that the y-axis is the beta value). 

*Note: the displayed table shows all results, however the MA plot does not show genes which did not pass filters, and Inf values.* 

```{r, sleuth_wald,  message=FALSE, fig.height=5, fig.width=6, warning=F}
so <- sleuth_wt(so, 'conditionKD')
results_table_wt <- sleuth_results(so, 'conditionKD', show_all = TRUE)

results_table_wt %>% datatable(options = list(pageLength=5))

p<-sleuth::plot_ma(so, "conditionKD", sig_color="green4", sig_level = 0.05)
p + theme_classic() + ggtitle("MA plot of genes and TEs, sleuth") + geom_text(data=(results_table_wt %>%  filter(str_detect(target_id,"(smt3|^R1_DM|^R2_DM)"))), aes(mean_obs, b, label=sub("_DM.*","",target_id)), position = position_nudge(x=0.5))

```

Fig. MA plot from sleuth analysis, Wald test. All genes with qvalue<0.05 are highlighted (warning: beta != log2FC)

Below are plots using the sleuth statistics, but including fold change information calculated from TPM values, and from scaled_reads_per_base_values. Combining four tables: srpb values, tmp values, and the LTR and wt statistics. 


```{r extended_tables}
### values tables
table.tpm<-sleuth_to_matrix(so, which_df = "obs_norm", which_units = "tpm") %>% as.data.frame() %>% rename(shW.1=YL25, shW.2=YL27, shSmt3.1=YL26, shSmt3.2=YL28) %>% tibble::rownames_to_column("target_id") %>% mutate(ctrl=(shW.1+shW.2)/2, kd=(shSmt3.1+shSmt3.2)/2) %>% mutate(log2FC=log2(kd/ctrl))

table.srbp<-sleuth_to_matrix(so, which_df = "obs_norm", which_units = units) %>% as.data.frame() %>% rename(shW.1=YL25, shW.2=YL27, shSmt3.1=YL26, shSmt3.2=YL28) %>% tibble::rownames_to_column("target_id") %>% mutate(ctrl=(shW.1+shW.2)/2, kd=(shSmt3.1+shSmt3.2)/2) %>% mutate(log2FC=log2(kd/ctrl))

table.values<-full_join(table.srbp, table.tpm, by="target_id",suffix=c(".srpb",".tpm"))
#NOTE: the logFC from the two values are very similar as expected

### stats tables
table.wald<-sleuth_results(so, 'conditionKD', test_type = 'wt', show_all = TRUE) %>% arrange(target_id) # wald test
table.lrt<-sleuth_results(so, 'reduced:full', test_type = 'lrt', show_all = TRUE) %>% arrange(target_id) #lrt test
table.stats<-full_join(table.wald, table.lrt, by="target_id", suffix=c(".wald",".lrt")) 

table.all<-full_join(table.stats, table.values, by="target_id") %>% arrange(desc(log2FC.tpm))
write.table(table.all, "output/table_all.txt", sep="\t", quote = F, col.names = T) # note that genes which didn't pass sleuth filters will have NA values instead of qval-s

```

Wald test MA plot, but I only marked genes with >2 fold change, calculated from the original values in the sleuth table above. Dots are colored by fold change increase
*Note: Some values that are Inf cannot be displayed.*

```{r, sleuth_wald_FC,  message=FALSE, fig.height=5, fig.width=8, fig.path='figures/', dev=c('png', 'pdf')}
tab.plot<-table.all %>% filter(!is.na(qval.wald)) %>% mutate(significant2FC=qval.wald<0.05 & abs(log2FC.srpb)>1) %>% mutate("type"=ifelse(TE.wald=="FALSE","RefSeq","TE"), "FC"=cut(2^abs(log2FC.srpb), breaks=c(0,2, 10, 100, Inf), right = F, include.lowest = T))

#tab.plot %>% ggplot(aes(mean_obs.wald, b)) + geom_point(aes(col=significant2FC), alpha=0.5) + scale_color_manual(values=c("black","green4"))

p<-tab.plot %>% ggplot(aes(mean_obs.wald, b)) + xlab(paste("mean( log(counts + 0.5 ) )")) + ylab(paste0("beta: conditionKD")) +  theme_classic() +  ggtitle("MA plot of genes and TEs, sleuth") + scale_shape_manual(values=c(16,15))
p<-p + geom_point(data=(tab.plot %>% filter(qval.wald>=0.05)), aes(mean_obs.wald, b, shape=type), size=2, alpha=0.5) 
p<-p + geom_point(data=(tab.plot %>% filter(qval.wald<0.05)), aes(mean_obs.wald, b, shape=type, color=FC), size=2, alpha=0.5) + scale_color_manual(values=c("black", plasma(10)[c(8,6,3)]))
p<-p + geom_text(data=(tab.plot %>% filter(str_detect(target_id,"(smt3|^R1_DM|^R2_DM)"))), aes(mean_obs.wald, b, label=sub("_DM.*","",target_id)), position = position_nudge(x=0.7))

p.box<-ggplot(tab.plot, aes(y=log2FC.srpb, x=type)) + geom_boxplot() + theme_light() + ggtitle("log2FC by type") + scale_y_continuous(breaks=seq(-5,10,1))

gridExtra::grid.arrange(p,p.box, widths=c(4,1))


```

Fig. MA plot from sleuth analysis, Wald test. (warning: beta != log2FC). Only genes with qval<0.05 & log2FC (from kallisto numbers) are highlighted.

***
***

The size factor of the wald test as Y-axis seems to be difficult to understand for broader reader. Plotting the same but:
1) shoing TPM values log2FC on the Y-axis, and mean between replicase on X-axis. Coloring will be according to the LRT (!) test.

```{r, FC_MA,  message=FALSE, fig.height=5, fig.width=8, fig.path='figures/', dev=c('png', 'pdf')}
tab.plot2<-table.all %>% filter(!is.na(qval.lrt)) %>% mutate(significant2FC=qval.lrt<0.05 & abs(log2FC.tpm)>1) %>% mutate("type"=ifelse(TE.wald=="FALSE","RefSeq","TE"), "FC"=cut(2^abs(log2FC.tpm), breaks=c(0,2, 10, 100, Inf), right = F, include.lowest = F)) #excluding infinite values which are un-interpretable

p<-tab.plot2 %>% ggplot(aes(log2((ctrl.tpm+kd.tpm)/2), log2FC.tpm)) + xlab(paste("log2 mean TPM (KD and control)")) + ylab("log2 FC TPM (KD vs control)") +  theme_classic() +  ggtitle("gene expression changes in smt3 KD") + scale_shape_manual(values=c(16,15))
p<-p + geom_point(data=(tab.plot2 %>% filter(qval.lrt>=0.05)), aes(log2((ctrl.tpm+kd.tpm)/2), log2FC.tpm, shape=type), size=2, alpha=0.5) 
p<-p + geom_point(data=(tab.plot2 %>% filter(qval.lrt<0.05)), aes(log2((ctrl.tpm+kd.tpm)/2), log2FC.tpm, shape=type, color=FC), size=2, alpha=0.5) + scale_color_manual(values=c("black", plasma(10)[c(8,6,3)]))
p<-p + geom_text(data=(tab.plot2 %>% filter(str_detect(target_id,"(smt3|^R1_DM|^R2_DM)"))), aes(log2((ctrl.tpm+kd.tpm)/2), log2FC.tpm, label=sub("_DM.*","",target_id)), position = position_nudge(x=0.7))

p.box<-ggplot(tab.plot2, aes(y=log2FC.tpm, x=type)) + geom_boxplot() + theme_light() + ggtitle("log2FC by type") + scale_y_continuous(breaks=seq(-5,10,1))

gridExtra::grid.arrange(p,p.box, widths=c(4,1))

```

***
***

### Alternative visualization

This is an alternative visualization of the data, where we do a scatter plot in control vs knockdown.
We can choose to select the dots based on **qvalues either** from the **Wald test**, or the **LTR test**. The test choice does not change the interpretation of the data in the light of the R1, R2 up-regulation compared to genes. We select the more stringent LTR test here.

```{r scatter, fig.width=15, fig.height=6, fig.path='figures/', dev=c('png', 'pdf')}
tab.plot3<-table.all %>% filter(!is.na(qval.lrt)) %>% mutate(significant2FC=qval.lrt<0.05 & abs(log2FC.tpm)>1) %>% mutate("type"=ifelse(TE.wald=="FALSE","RefSeq","TE"), "FC"=cut(2^abs(log2FC.tpm), breaks=c(0,2, 10, 100, Inf), right = F, include.lowest = T)) # no need to exclude infinites here, as this is just for coloring

top20.x<-tail(sort(tab.plot3$ctrl.tpm),20)[1]
top20.y<-tail(sort(tab.plot3$kd.tpm),20)[1]

p<-ggplot(tab.plot3, aes(log2(ctrl.tpm+1), log2(kd.tpm+1))) + theme_classic() + theme(legend.position="left") + xlab(paste("shW, log2 (av.tpm + 1) ")) + ylab(paste("shSmt3, log2 (av.tpm + 1) ")) + geom_hline(yintercept = log2(top20.y+1), linetype="dotted") + geom_vline(xintercept = log2(top20.x+1), linetype="dotted") + ggtitle("gene and TE expression, RNA-seq [TPM]") + scale_shape_manual(values=c(16,15))
p<-p + geom_point(data=(tab.plot3 %>% filter(qval.lrt>=0.05)), aes(log2(ctrl.tpm+1), log2(kd.tpm+1), shape=type), size=2, alpha=0.5) 
p<-p + geom_point(data=(tab.plot3 %>% filter(qval.lrt<0.05)), aes(log2(ctrl.tpm+1), log2(kd.tpm+1), shape=type, color=FC), size=2, alpha=0.5) + scale_color_manual(values=c("black", plasma(10)[c(8,6,3)]))
p<-p +  geom_text(data=(tab.plot3 %>% filter(str_detect(target_id,"(smt3|^R1_DM|^R2_DM)"))), aes(log2(ctrl.tpm+1), log2(kd.tpm+1), label=sub("_DM.*","",target_id)), position = position_nudge(x=0.7)) + geom_abline(intercept = 0, slope=1, col="blue")  

p.hist<-ggExtra::ggMarginal(p, type="histogram", size=3, fill = "grey70")

gridExtra::grid.arrange(p, p.hist, nrow=1)

```

Figure. Scatter plots of gene expression in control vs KD. The dotted line denotes area that contains the top 20 most highly expressed genes in the corresponding condition. Values log2FC>2 and significant: qvalue (LRT) < 0.05 are highlighted.
The histograms above the y-axis and x-axis show the distribution of values.

*Note: Here, pale dots show values ("NA") which did not pass various sleuth filters and should be ignored. Can be removed by "show_all=FALSE".*


***
***


## Numerical summary


Number of RepBase (TE) and other (protein-coding genes,ncRNA, etc) entries from RefSeq (all genes) analyzed; some were filtered out because they didn't pass sleuth criteria (usually very low read counts or large discrepancy between reps).

```{r num_summary}
table.all<-table.all %>% mutate("type"=ifelse(TE.wald=="FALSE","RefSeq","TE"), filt_out=is.na(b)) 
table.all %>% group_by(type, filt_out) %>% tally() %>% arrange(filt_out)
```

***

Number of significantly changed genes (qval 0.05, LRT or Walt Test; logFC (onscaled_reads_per_base)):

```{r signFC_by_type}
#Wald test
table.all %>% filter(abs(log2FC.tpm)>1, qval.wald<0.05) %>% mutate("change"=ifelse(log2FC.tpm>0,"up", "down")) %>% group_by(type,change) %>% tally() %>% arrange(desc(change))
#LRT
table.all %>% filter(abs(log2FC.tpm)>1, qval.lrt<0.05) %>% mutate("change"=ifelse(log2FC.tpm>0,"up", "down")) %>% group_by(type,change) %>% tally() %>% arrange(desc(change))

```

#### Summary of the numbers above:
From 8996 RefSeq genes and 166 RebBase entries which passed the initial filters, 
numbers indicated above are significantly up-regulated or down-regulated more than 2-fold in smt3 KD, (qvalue 0.05, Wald test or LTR test).
Note that if we calculate the fold change based on scaled_reads_per_base we get the same results with differences in only 1-2 genes.

**Therefore, smt3 KD led to >2 fold up-regulation for ~30% of the TE-s and ~4% of the annotated genes (protein-coding and ncRNAs), and down-regulation of ~1.5% of the annotated genes which passed initial filters. Without filtering, the percentages are around 2% and 0.8% for genes, and 21% for TEs** 

**R1 and R2 - from relatively lowly expressed genes in the control - become among the top 20 most abundant products in SUMO KD.**  : see below
***

Although several hundred genes change, it is evident from the plots mostly modestly expressed genes that show the large FC, except for R1 and R2. If we filter genes with over 100 times change in expression log2FC>=~6.64, significant in at least one test:

```{r FC_filter, warning=FALSE, message=F}
table.all %>% filter(log2FC.tpm>log2(100)) %>% filter(qval.wald<0.05 | qval.lrt<0.05) %>%
                    select(target_id, shW.1.tpm, shW.2.tpm, shSmt3.1.tpm, shSmt3.2.tpm, log2FC.tpm, shW.1.srpb, shW.2.srpb, shSmt3.1.srpb, shSmt3.2.srpb, log2FC.srpb, type, b, qval.wald, qval.lrt) %>% mutate_at(2:5, funs(round(.,3))) %>% arrange(-( shSmt3.1.srpb*0.5+shSmt3.2.srpb*0.5)) %>% datatable(options=list(pageLength=5))
```

Results show that all other genes, apart from R1 and R2 in this set, are ones that practically have no reads in the control, and just few reads in the KD. 

***


Numbers below show the xpression of R1 and R2 elements; including rank of the average values. "X" and "Y" are average per control and KD, respectively:

*Note: Values can be seen as transcripts per million (TPM), see https://haroldpimentel.wordpress.com/2014/05/08/what-the-fpkm-a-review-rna-seq-expression-units/
, or scaled_reads_per_base (gene mode only): The average number of reads mapping to each base across the whole gene*

```{r Rs_expr_rank}
tab.all.ranks1<-table.all %>% filter(!filt_out) %>% select(target_id, matches("sh.*srpb"),matches("(ctrl|kd).srpb")) %>% mutate(ctrl.rank.srpb=rank(desc(ctrl.srpb)), kd.rank.srpb=rank(desc(kd.srpb)))

tab.all.ranks2<-table.all %>% filter(!filt_out) %>% select(target_id, matches("sh.*tpm"),matches("(ctrl|kd).tpm")) %>% mutate(ctrl.rank.tpm=rank(desc(ctrl.tpm)), kd.rank.tpm=rank(desc(kd.tpm)))

tab.all.ranks1 %>% filter(str_detect(target_id, "^R[12]_"))
tab.all.ranks2 %>% filter(str_detect(target_id, "^R[12]_"))
```

**Notably, R1 and R2 change from relatively lowly expressed elements, to the top 20 most abundant transcripts (in TPM). No other elements shows such a dramatic increase.**

***
I'd like to thank Lynn Yi for the greatly helpful tips on using kallisto and sleuth.